Historical trends in the contribution of six major causes of death to the black-white

### 2017-04-26 Manuscript/Manuscript_COD_Trends.Rmd
bib changes

To calculate life expectancy at birth, the smoothed death counts were aggregated to estimate the total deaths in each age group and standard life table methods were applied [@selvin_text; @preston_demography_text]. Life expectancy at birth (in years) was calculated for blacks and whites in each state and year according to sex. The difference in life expectancy between blacks and whites was calculated by subtracting the black life expectancy from the white life expectancy in each state, year, and sex.

To examine the components of the difference in life expectancy between blacks and whites, we performed a decomposition analysis using the Arriaga method [ref]. First, the model partitions the difference in life expectancy by age group, which can be interpreted as the amount of the gap (in years) that can be attributed to higher age-specific mortality rates in blacks. For example, a decomposition of a gap of five years may find that two of the five years (40% of the gap) is attributable to a higher mortality rate in blacks in the youngest age group (< 1 years old) compared with the youngest whites. This age-specific decomposition can be further partitioned according to cause of death, producing the number of years of the gap that is attributable to every age- and cause-of-death combination. For example, a decomposition of a gap of 8 years may find that 1.3 years of this gap is attributable to more injury-related deaths in blacks aged 25-29 years old. We summed these age- and cause-of-death partitions over all ages to calculate the marginal contribution of each cause-of-death to the difference, for each state and for each sex. Thus, for each state, the contribution of each of the six causes to the black-white life expectancy gap was estimated for each year, producing 45 state-level trends and their credible intervals.

We also calculated age-standardized mortality rates for year-state-race-sex stratum, to aid the interpretation of the decomposition results. We used the US standard population from the year 2000 [ref].

@@

**Figures 1 and 2** display trends in the life expectancy gap for males and females, respectively. For both sexes, the life expectancy gap has decreased since 1969, although several states exhibited a period of increasing inequality that peaked in the early 1990s, especially among males. *Don't know whether you want to note it here or not, but I find it is interesting to see that the gap is now lower in many US southern (or southeastern) states than in the Midwest, suggesting that progress has actually been somewhat greater in the South than elsewhere (particularly the upper Midwest and especially for women). I think this is interesting and somewhat counter to the narrative that health indicators are worse in the South. I also wonder how this might relate to the Murray/IHME work on US counties with "declining" life expectancy among white women.*

@@

Injury is one of the least common causes of death. Over time, injury-related mortality has decreased precipitously among blacks, although an [insurgence][better word?] of crime resulted in heightened mortality from injuries among black males during the 1990s (**eFigure X**). For whites, most states exhibit an initial period of gradual decreases in injury mortality, followed by a period of gradual increases, especially among white females (**eFigure X**). Thus, injury mortality rates among blacks are the same as, or lower than those among whites in many states, with white females demonstrating sharper increases. Thus, the contribution of injuries to the mortality gap has generally decreased over time, and has become negative in those states with lower injury-related mortality among blacks than whites (**Figures 9 and 10**). Among males, DC, Illinois, Michigan and Missouri exhibit the largest recent injury-related contributions. Among females, many southern states exhibit negative contributions (indicating lower injury-related mortality among black women), while DC, Illinois, and Michigan having higher contributions than the general trend.

@@

In this paper, we examined state-level trends in six major CODs' contributions to the black-white life expectancy gap. While we found that many states mimicked the national patterns of cause-of-death specific contributions, several states diverged in ways that either resulted in the near elimination of the life expectancy gap, or a less favorable result. We found that cardiovascular disease contributes the most to the life expectancy gap, even though both races and genders experienced decreased CVD mortality rates over time. Injuries continue to play a role in contributing to the current gap for males in several states, especially in the Midwest and DC, while contributing to a decreased gap among females in the South. Illinois, Michigan and Wisconsin also show relatively high contributions from cancer. On the other hand, New York and Massachusetts have virtually eliminated the black-white life expectancy gap for both men and women. However, both states began with only a minimal contribution of CVD and cancer to the mortality gap. The difference in early contributions is especially stark for the contribution of CVD among women in 1969, where many states, especially those in the South, had CVD contribution of more than two years. Thus, the reason for New York’s and Massachusetts’s low CVD contribution does not appear to be the effect of recent changes, but highlights a long-standing minimal difference and point to the importance of viewing the most recent estimates in light of their earlier values.

@@

This research extends previous studies that have estimated either the contribution of causes of death to the life expectancy gap on the national level, or the excess risk among blacks in cause of death specific mortality rates (references: Harper 2007, 2012; kochanek 2013, Orsi 2010).

@@

census_map: https://www.census.gov/2010census/popmap/. retrived apr 25, 2017

>>>>

To calculate life expectancy at birth, the smoothed death counts were aggregated to estimate the total deaths in each age group and standard life table methods were applied [@selvin_2004; @preston]. Life expectancy at birth (in years) was calculated for blacks and whites in each state and year according to sex. The difference in life expectancy between blacks and whites was calculated by subtracting the black life expectancy from the white life expectancy in each state, year, and sex.

To examine the components of the difference in life expectancy between blacks and whites, we performed a decomposition analysis by cause of death using the Arriaga method [@arriaga1984measuring]. This method partitions the black-white difference in life expectancy by cause of death, which can be interpreted as the amount of the gap, in years, that can be attributed to higher age-standardized cause-specific mortality rates among blacks. For example, a decomposition of a gap of five years may find that two of the five years is attributable to a higher age-standardized rate of mortality due to cardiovascular disease among blacks compared to whites. If the decomposition produces a negative estimate, say -0.5 years for injuries, this implies that whites have a higher age-standardized rate of mortality for injuries compared to blacks. Thus, for each state, the contribution of each of the six causes to the black-white life expectancy gap was estimated for each year, producing 45 state-level trends and their credible intervals.

We also calculated age-standardized mortality rates for year-state-race-sex stratum, to aid the interpretation of the decomposition results. We used the US standard population from the year 2000 [ref census_stdpop].

@@

**Figures 1 and 2** display trends in the life expectancy gap for males and females, respectively. For both sexes, the life expectancy gap has decreased since 1969, although several states exhibited a period of increasing inequality that peaked in the early 1990s, especially among males.

*SH says: Don't know whether you want to note it here or not, but I find it is interesting to see that the gap is now lower in many US southern (or southeastern) states than in the Midwest, suggesting that progress has actually been somewhat greater in the South than elsewhere (particularly the upper Midwest and especially for women). I think this is interesting and somewhat counter to the narrative that health indicators are worse in the South. I also wonder how this might relate to the Murray/IHME work on US counties with "declining" life expectancy among white women.*

@@

Injury is one of the least common causes of death. Over time, injury-related mortality has decreased precipitously among blacks, although an [insurgence][*better word?*] of crime resulted in heightened mortality from injuries among black males during the 1990s (**eFigure X**). For whites, most states exhibit an initial period of gradual decreases in injury mortality, followed by a period of gradual increases, especially among white females (**eFigure X**). Thus, injury mortality rates among blacks are the same as, or lower than those among whites in many states, with white females demonstrating sharper increases. Thus, the contribution of injuries to the mortality gap has generally decreased over time, and has become negative in those states with lower injury-related mortality among blacks than whites (**Figures 9 and 10**). Among males, DC, Illinois, Michigan and Missouri exhibit the largest recent injury-related contributions. Among females, many southern states exhibit negative contributions (indicating lower injury-related mortality among black women), while DC, Illinois, and Michigan having higher contributions than the general trend.

@@

In this paper, we examined state-level trends in six major CODs' contributions to the black-white life expectancy gap. While we found that many states mimicked the national patterns of cause-of-death specific contributions, several states diverged in ways that either resulted in the near elimination of the life expectancy gap, or a less favorable result.

We found that cardiovascular disease contributes the most to the life expectancy gap, even though both races and genders experienced decreased CVD mortality rates over time. Injuries continue to play a role in contributing to the current gap for males in several states, especially in the Midwest and DC, while contributing to a decreased gap among females in the South. Illinois, Michigan and Wisconsin also show relatively high contributions from cancer. On the other hand, New York and Massachusetts have virtually eliminated the black-white life expectancy gap for both men and women. However, both states began with only a minimal contribution of CVD and cancer to the mortality gap. The difference in early contributions is especially stark for the contribution of CVD among women in 1969, where many states, especially those in the South, had CVD contribution of more than two years. Thus, the reason for New York’s and Massachusetts’s low CVD contribution does not appear to be the effect of recent changes, but highlights a long-standing minimal difference and point to the importance of viewing the most recent estimates in light of their earlier values.

@@

This research extends previous studies that have estimated either the contribution of causes of death to the life expectancy gap on the national level, or the excess risk among blacks in cause of death specific mortality rates [@harper2007trends; @harper2012trends; orsi2010black] [ref kochanek_2013].

@@

census_map: https://www.census.gov/2010census/popmap/. retrieved apr 25, 2017

census_std_pop: https://seer.cancer.gov/stdpopulations/stdpop.19ages.html. retrieved apr 25, 2017

kochanek_2013: https://www.cdc.gov/nchs/data/databriefs/db125.pdf
